Dynamic Stochastic Approximation Monte Carlo

### 2018-05-16 papers/histogram/sad.tex
start adding a figure comparing the entropy for N500
adding text to paper sad.

\section{SAD Algorithm}
\subsubsection{The theory behind SAD}
The SAD (Dynamic Stochastic Approximation) is a version of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. While for
SAMC the update factor is defined in the original
implementation~\cite{liang2007stochastic} in terms of an arbitrary
tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} = \frac{t_0}{\max(t_0,t)}
\end{align}
The update factor $\gamma_{t}^{\text{SA}}$ can be thought of as
$\nicefrac{\text{d}S}{\text{d}t}$. This tells us that $t_0$ should have
dimensions of entropy.  We argue that we can write this entropy `term'
geometrically as an area bounded by the $\ln{D(E)}$ i.e. the entropy.  This
gives the following
\begin{align}
\gamma_{t}^{\text{SA}} =
\frac{N_E}{t}\frac{E_{\text{max}}-E_{\text{min}}}{3T_{\text{min}}}
\end{align}
where $N_E$ is the number of energies that have been found. The change
in the entropy the minimum temperature can be thought of as a slope of
the triangle made by connecting the minimum important energy and the
maximum entropy state.  The factor of $3$ comes from the ratio of areas bounded
by the $\ln{D(E)}$ and the minimum temperature line.
The update factor has the same $1/t$ behavior as the original SAMC algorithm;
however now every time a new energy is discovered during the course
of the simulation, the update factor experiences a jump. This behavior allows
SAD to dynamically prevent the update factor from decreasing too rapidly.

Fig.~\ref{fig:gamma-vs-t}. shows the monotonic decrease of the update
factor for SAMC.  The update factor for WL decreases like a step
function while the update factor for SAD dynamically fluctuates around
a value less than 1 for early MC moves (while finding new energy
states).  Ultimately once all energy states have been found, the update
factor for SAD decreases in the same monotonic fashion as SAMC.

\subsubsection{High and Low energy racheting}
Since SAD explores all energy states it is imperative that the method have
a reasonable idea about where the minimum important energy and maximum
entropy state are located in energy space. Since these are not user input
parameters like for methods suchas WL-TMMC, the simulation is responsible
for determining and updating these energy locations. In order to prevent,
the method from experiencing errors near the energy extrema, we employ a 
high energy and a low energy to define the range over which the energy 
histogram is made flat. The high energy is the highest ever value that the
maximum entropy state has taken while the low energy is the lowest ever value
that the minimum important energy has taken. Together these two energies serve
as a kind of ratcheting effect to determine the energy landscape boundaries.

>>>>

\section{SAD Algorithm}
\subsubsection{The theory behind SAD}
~\jpsays{We need to define or talk about Tmin.  This is a disadvantage
of of SAD. WL requires an energy range. SAMC removes the energy range
requirement but requires simulating every possible energy.  For SAD,
Tmax set to $\infty$ is fine.  Easy to simulate this way. SAD focuses
on energies between Tmin and infinity.}

The SAD (Dynamic Stochastic Approximation) is a version of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. While for
SAMC the update factor is defined in the original
implementation~\cite{liang2007stochastic} in terms of an arbitrary
tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} = \frac{t_0}{\max(t_0,t)}
\end{align}
The update factor $\gamma_{t}^{\text{SA}}$ can be thought of as
$\nicefrac{\text{d}S}{\text{d}t}$. This tells us that $t_0$ should have
dimensions of entropy.  We argue that we can write this entropy `term'
geometrically as an area bounded by the $\ln{D(E)}$ i.e. the entropy.  This
gives the following
\begin{align}
\gamma_{t}^{\text{SA}} =
\frac{N_E}{t}\frac{E_{\text{max}}-E_{\text{min}}}{3T_{\text{min}}}
\end{align}
where $N_E$ is the number of energies that have been found. The change
in the entropy the minimum temperature can be thought of as a slope of
the triangle made by connecting the minimum important energy and the
maximum entropy state.  The factor of $3$ comes from the ratio of areas
bounded by the $\ln{D(E)}$ and the minimum temperature line. The update
factor has the same $1/t$ behavior as the original SAMC algorithm;
however now every time a new energy is discovered during the course of
the simulation, the update factor experiences a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly. In contrast to SAD, Fig.~\ref{fig:gamma-vs-t}. shows the
monotonic decrease of the update factor for SAMC.  The update factor
for WL decreases like a step function while the update factor for SAD
dynamically fluctuates around a value less than 1 for early MC moves
(while finding new energy states).  Ultimately once all energy states
have been found, the update factor for SAD decreases in the same
monotonic fashion as SAMC.

\subsubsection{High and Low energy racheting}
~\jpsays{change name to $\epsilon_H$ and $\epsilon_L$. Include formula
for updating weights above and below high/low energy.  Explain how we
compute the maximum entropy state and minimum important energy.  Then
we are in a position to talk about racheting.} Since SAD explores all
energy states it is imperative that the method have a reasonable idea
about where the minimum important energy and maximum entropy state are
located in energy space. Since these are not user input parameters like
for methods suchas WL-TMMC, the simulation is responsible for
determining and updating these energy locations. In order to prevent,
the method from experiencing errors near the energy extrema, we employ
a high energy $\epsilon_H$ and a low energy $\epsilon_L$ to define the
range over which the energy histogram is made flat. $\epsilon_H$ is the
highest ever value that the maximum entropy state has taken while
$\epsilon_L$ is the lowest ever value that the minimum important energy
has taken. SAD uses these two energies to control the weights $w(E)$
above $\epsilon_H$ and $\epsilon_L$. If the energy $E$ is less than
$\epsilon_H$ ~\jpsays{greater than sign backward} then the weights can
be calculated as ~\jpsays{equation formatting problems!}
\begin{equation}
\begin{aligned}
\ln w(E) = \min(\ln w(E) - \gamma, \\
    \ln w(\epsilon_H) - \log(e^{\ln w(\epsilon_H) -w(E)}))
\end{aligned}
\end{equation}
If the energy $E$ is greater than $\epsilon_L$ ~\jpsays{greater than
sign backward} then the weights can be calculated as
\begin{equation}
\begin{aligned}
\ln w(E) = \min(\ln w(E) - \gamma, \ln w(\epsilon_L) \\
    + \frac{E-\epsilon_L}{T_\text{min}}
    - \log(\gamma + e^{\ln w(\epsilon_L) -w(E)
    + \frac{E-\epsilon_L}{T_\text{min}}}))
\end{aligned}
\end{equation}
Sad sets the maximum entropy state equal to the energy if either of the
following is true.~\jpsays{less than sign backward}
\begin{align}
\ln w(E) < \ln w(S_\text{max}) \; \text{or} \; H(S_\text{max}) = 0
\end{align}
Thus if the method is beyond the maximum entropy state or has collected no
statistics at the maximum entropy state then it is set equal to the energy.
Similarly, Sad sets the minimum important energy equal to the energy if
either of the following is true.~\jpsays{check less than sign}
~\jpsays{equation formatting problems!}
\begin{align}
\ln w(E) < \ln w(E_\text{min}) + \frac{E-E_\text{min}}{T_\text{min}} 
\; \text{or} \; H(E_\text{min}) = 0
\end{align}
Thus if the method is beyond the minimum important energy or has collected no
statistics at the minimum important energy then it is set equal to the energy.

Together these two
energies serve as a kind of ratcheting effect to determine the energy
landscape boundaries.
~\jpsays{So outside the range set by the weights $w(E)$, the maximum entropy
state is set to the energy $E$. Likewise with minimum important energy.}
